A full approximation scheme multilevel method \\ for nonlinear variational inequalities

### 2023-07-22 paper/fascd.tex
clean abstract and introduction; address a PEF

\begin{abstract}
We present the \emph{full approximation scheme constraint decomposition} (FASCD) multilevel method for solving variational inequalities (VIs).  FASCD is a common extension both of the full approximation scheme (FAS) multigrid technique for nonlinear partial differential equations, due to A.~Brandt, and the constraint decomposition (CD) method introduced by X.-C.~Tai for VIs arising in optimization.  We extend the CD idea by exploiting the telescoping nature of certain function space subset decompositions arising from multilevel mesh hierarchies.  We demonstrate that FASCD V-cycles exhibit excellent mesh-independent convergence rates and that FASCD F-cycles (full multigrid) are optimal solvers, when used in combination with reduced-space active set smoothers with work proportional to the number of unknowns on a given mesh level.  The example problems include unilateral and bilateral VI problems, with differential operators which are symmetric linear, nonsymmetric linear, and nonlinear.  Our examples demonstrate very good convergence even on problems where other multilevel VI solvers struggle.
\end{abstract}

@@

The constraint decomposition (CD) methods of Tai \cite{Tai2003} are designed to solve coercive variational inequality (VI) problems arising as the Karush--Kuhn--Tucker optimality conditions for the minimization of a convex functional over a convex set.  These methods were shown to converge for multilevel and domain decompositions in a finite element (FE) context; in the multilevel case, the CD method has almost grid-independent error bounds, and near-optimal complexity, for elliptic, linear obstacle problems \cite[Subsection 5.4]{Tai2003}, \cite[Theorem 4.6 and Algorithm 4.7]{GraeserKornhuber2009}.
In this paper we extend Tai's multilevel CD method by replacing the coarser-level corrections with a full approximation scheme (FAS; \cite{Brandt1977,Bruneetal2015}) approach.  Specifically, our \emph{full approximation scheme constraint decomposition} (FASCD) method extends the algorithms in \cite{GraeserKornhuber2009} and \cite{Tai2003} in the following five directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item The FAS coarse corrections are adapted for general VI problems. The needed computations require neither an objective function nor linearity of the residual operator.
\item We identify a ``telescoping'' subset property which can be exploited on the up-smoothing side of a V-cycle for greater performance.
\item Box constraints, namely upper and lower obstacles, are permitted.
\item We implement V-cycles (Algorithm \ref{alg:fascd}) and F-cycles (full multigrid; Algorithm \ref{alg:fascd-fcycle}).
\item We demonstrate that F-cycle FASCD is an optimal solver for the different VI examples in Section \ref{sec:results}, when used in conjunction with reduced-space active set methods as smoothers. The number of Newton and Krylov iterations per step are fixed, so that the work of the smoother is proportional to the number of unknowns on a given level.
\pef{The solver works for VIs where the residual is only defined on the admissible set \dots}
\end{enumerate}

Prior multilevel algorithms for VIs have various substantial limitations compared to the FASCD framework.  The projected full approximation scheme (PFAS) algorithm of Brandt \& Cryer \cite{BrandtCryer1983} is apparently limited to linear complementarity problems and projected (truncated) Gauss--Seidel smoothers.  While FASCD reduces to general, nonlinear FAS for PDEs when the inequality constraints are removed (Appendix \ref{app:reductions}), PFAS reduces to linear multigrid.
The multilevel CD iteration of Tai \cite{Tai2003} is formulated in terms of an objective function, so as stated it only applies to optimization-type problems, and furthermore each of its coarser-level problems refers to the finest-level discretization.  The problematic latter property is corrected by reformulation using defect constraints and monotone injection operators (see \cite[Algorithm 4.7]{GraeserKornhuber2009} and Section \ref{sec:cdmultilevel} herein).  The monotone multigrid methods of Kornhuber and co-authors \cite{GraeserKornhuber2009,Kornhuber1994} appear to be formulated only with Gauss--Seidel type smoothers, and applied only to optimization-type VI problems \cite[for example]{JouvetGraeser2013}.  Furthermore, monotone multigrid methods are more difficult to implement in FE libraries because modifications of the FE basis functions are needed to approximate the coarser-level coincidence sets.  No such invasive basis modifications are used either in Tai's original multilevel CD or in FASCD.  Our implementation of FASCD in the Firedrake library \cite{Rathgeberetal2016} requires no modifications of basis functions and yet exhibits excellent performance, including on a problem used to demonstrate the benefits of monotone multigrid (Example \ref{ex:results:classical}; compare \cite[problem 7.1.1]{GraeserKornhuber2009}).

In our experiments, FASCD V-cycles exhibit essentially the same mesh-independent convergence rates for VI problems that FAS multigrid V-cycles exhibit for the corresponding (unconstrained) PDE problems.  The F-cycle results in Section \ref{sec:results} show nearly-textbook multigrid efficiency, namely solutions within discretization error in work comparable to few smoother sweeps on the finest-level mesh \cite{BrandtLivne2011}.  Specifically, we see this performance for classical Laplacian obstacle problems, $p$-Laplacian obstacle problems, and nonnegativity-constrained advection-diffusion equations.  Furthermore, these results appear to be insensitive to the length or structure of the free boundary, i.e.~the geometric complexity of the coincidence set.

All iterates in FASCD are admissible, and thus the nonlinear operator in the VI needs only be defined for admissible states.  For example, we solve an ice sheet geometry problem (Examples \ref{ex:sia} and \ref{ex:results:sia}) in which the ice flow formulas are only meaningful for surface elevation iterates that do not penetrate the bedrock.  (More generally, fluid layers subject to a non-negative thickness constraint typically require the solution of an auxiliary PDE on a domain determined inside the VI residual evaluation \cite{Bueler2021conservation}, but this is not needed in the ice sheet problem here.)  Non-admissible methods, including semi-smooth methods \cite{BensonMunson2006}, would require unnatural modifications of the operator formula.  FASCD solves this low-regularity ice sheet problem, wherein the solution has unbounded gradient at the free boundary, by a mesh-independent number of F-cycle iterations (Example \ref{ex:results:sia}).

The paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone variational inequalities, and describe several motivating example problems.  Section \ref{sec:cd} reviews the basics of CDs and their associated iterations.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, and then Section \ref{sec:cdmultilevel} presents defect constraints generated via monotone injection operators \cite{GraeserKornhuber2009}.  Here we propose an apparently new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD and an associated iteration.  Section \ref{sec:vcycle} then states the FASCD V-cycle, with Section \ref{sec:implementation} addressing convergence criteria, F-cycles, and smoother construction.  Section \ref{sec:results} gives numerical results, from a parallel Firedrake \cite{Rathgeberetal2016} implementation (see Code Availability), for each example problem introduced in Section \ref{sec:vi}.  The work concludes with a discussion and outlook in Section \ref{sec:discussion}.

>>>>

\begin{abstract}
We present the \emph{full approximation scheme constraint decomposition} (FASCD) multilevel method for solving variational inequalities (VIs).  FASCD is a common extension both of the full approximation scheme (FAS) multigrid technique for nonlinear partial differential equations, due to A.~Brandt, and the constraint decomposition (CD) method introduced by X.-C.~Tai for VIs arising in optimization.  We extend the CD idea by exploiting the telescoping nature of certain function space subset decompositions arising from multilevel mesh hierarchies.  We demonstrate that FASCD V-cycles exhibit excellent mesh-independent convergence rates.  Furthermore, when a reduced-space active set method is applied as a smoother, with work proportional to the number of unknowns on a given mesh level, FASCD F-cycles (full multigrid) are optimal solvers.  The example problems include differential operators which are symmetric linear, nonsymmetric linear, and nonlinear, in unilateral and bilateral VI problems.  Our examples demonstrate very good convergence even on problems where other multilevel VI solvers struggle.
\end{abstract}

@@

The constraint decomposition (CD) methods of Tai \cite{Tai2003} are designed to solve coercive variational inequality (VI) problems arising as the Karush--Kuhn--Tucker optimality conditions for the minimization of a convex functional over a convex set.  These methods were shown to converge for multilevel and domain decompositions in a finite element (FE) context.  In the multilevel case, this CD method has almost grid-independent error bounds, and near-optimal complexity, for elliptic, linear obstacle problems \cite[Subsection 5.4]{Tai2003}, \cite[Theorem 4.6 and Algorithm 4.7]{GraeserKornhuber2009}.

Here we extend Tai's multilevel CD method by replacing the coarser-level corrections with a full approximation scheme (FAS; \cite{Brandt1977,Bruneetal2015}) approach.  Specifically, our \emph{full approximation scheme constraint decomposition} (FASCD) method extends the \cite{GraeserKornhuber2009,Tai2003} algorithms in the following directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item The FAS coarse corrections are adapted for general VI problems. The needed computations require neither an objective function nor linearity of the residual operator.
\item Box constraints, namely arbitrary upper and lower obstacles, are allowed.
\item As with the original CD method \cite{Tai2003}, but unlike some implementations \cite[for example]{GraeserKornhuber2009}, and semismooth methods generally \cite{BensonMunson2006,Ulbrich2011}, the method can be applied even if the residual operator is defined only on those functions which satisfy the constraints.
\item V-cycles (Algorithm \ref{alg:fascd}) and F-cycles (full multigrid; Algorithm \ref{alg:fascd-fcycle}) are implemented.
\item We identify a ``telescoping'' subset property which can be exploited on the up-smoothing side of a V-cycle for greater performance.
\item When used in conjunction with a reduced-space active set method as a smoother, we demonstrate that F-cycle FASCD is an optimal solver for the different VI examples in Section \ref{sec:results}.  The work of the smoother is proportional to the number of unknowns on a given level because the number of Newton and Krylov iterations is fixed.
\end{enumerate}

Prior multilevel algorithms for VIs have various substantial limitations compared to the FASCD framework.  The projected full approximation scheme (PFAS) algorithm of Brandt \& Cryer \cite{BrandtCryer1983} is apparently limited to linear complementarity problems and projected (truncated) Gauss--Seidel smoothers.  While FASCD reduces to general, nonlinear FAS for PDEs when the inequality constraints are removed (Appendix \ref{app:reductions}), PFAS reduces to linear multigrid.  The multilevel CD iteration of Tai \cite{Tai2003} is formulated in terms of an objective function, so it only applies to optimization-type problems, and furthermore each of its coarser-level problems refers to the finest-level discretization.  The latter property is improved by reformulation using defect constraints and monotone injection operators (see \cite[Algorithm 4.7]{GraeserKornhuber2009}, or Section \ref{sec:cdmultilevel} herein).  The monotone multigrid methods of Kornhuber and co-authors \cite{GraeserKornhuber2009,Kornhuber1994} appear to be formulated only with Gauss--Seidel type smoothers, and applied only to optimization-type VI problems \cite[for example]{JouvetGraeser2013}.  Furthermore, monotone multigrid methods are more difficult to implement in FE libraries because modifications of the FE basis functions are needed to approximate the coarser-level coincidence sets.  No such invasive basis modifications are used either in Tai's CD, nor in our Firedrake library \cite{Rathgeberetal2016} implementation of FASCD.  However, our implementation exceeds monotone multigrid performance, including on a problem used to demonstrate the benefits of that method (Example \ref{ex:results:classical}; compare \cite[problem 7.1.1]{GraeserKornhuber2009}).

Regarding its limitations, FASCD solves only bound-constrained VI problems.  The method differs from the original FAS method only if the constraint set is proper (Appendix \ref{app:reductions}).

In our experiments, FASCD V-cycles exhibit essentially the same mesh-independent convergence rates for VI problems that FAS multigrid V-cycles exhibit for corresponding unconstrained PDE problems.  The F-cycle results in Section \ref{sec:results} show nearly-textbook multigrid efficiency, namely solutions within discretization error in work comparable to few smoother sweeps on the finest-level mesh \cite{BrandtLivne2011}.  Specifically, we see this performance for classical (unilateral, Laplacian) obstacle problems, $p$-Laplacian obstacle problems, and box-constrained advection-diffusion equations.  Furthermore, these results appear to be insensitive to the length or structure of the free boundary, i.e.~the geometric complexity of the coincidence set.

All iterates in FASCD are admissible, and thus the nonlinear operator in the VI needs only be defined for admissible states.  For example, we solve an ice sheet geometry problem (Examples \ref{ex:sia} and \ref{ex:results:sia}) in which the ice flow formulas are only meaningful for surface elevation iterates that do not penetrate the bedrock.  Non-admissible methods, including semi-smooth methods \cite{BensonMunson2006}, would require unnatural modifications of the operator formula.  FASCD solves this low-regularity ice sheet problem, wherein the solution has unbounded gradient at the free boundary, by a small, mesh-independent number of F-cycle iterations (Example \ref{ex:results:sia}).

{\color{blue} ELB: beam 2d example?}

The paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone variational inequalities, and describe several motivating examples.  Section \ref{sec:cd} reviews the basics of CDs and their associated iterations.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, and then Section \ref{sec:cdmultilevel} extends the CD method via defect constraints generated by monotone injection operators \cite{GraeserKornhuber2009}.  Here we propose an apparently new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD and an associated iteration.  Section \ref{sec:vcycle} then states the FASCD V-cycle, with Section \ref{sec:implementation} addressing convergence criteria, F-cycles, and our smoother choice.  Section \ref{sec:results} gives numerical results from a parallel Firedrake \cite{Rathgeberetal2016} implementation (see Code Availability), for each example problem introduced in Section \ref{sec:vi}.  The work concludes with a discussion and outlook in Section \ref{sec:discussion}.
